---
title: "Data Analysis with R, Python and Rcpp"
output: 
  html_document: 
    toc: yes
    fig_caption: yes
    number_sections: yes
date: "2023-09-27"
---

-   **Contributor:** P.BADY
-   **Date:** `r date()`
-   **Abstract:**


```         
License: GPL version 2 or newer
Copyright (C) 2009-2017  Pierre Bady

This program is free software/document; you can redistribute it and/or
modify it under the terms of the GNU General Public License
as published by the Free Software Foundation; either version 2
of the License, or (at your option) any later version.

This program/document is distributed in the hope that it will be useful,
but WITHOUT ANY WARRANTY; without even the implied warranty of
MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
GNU General Public License for more details.
```

# Motivations
The objective of this document is to propose an implementation of data analysis methods based on duality diagram in python and to develop and expand my skill in python. The document contains information related to the use of Rcpp (R and C++).


# first test PCA based on R package ade4
<a href="#top">Back to top</a>

***

## Example from the R package ade4

```{r}
library(ade4)
data(deug)
pca1 <- dudi.pca(deug$tab, center = deug$cent, scale = FALSE, scan = FALSE)
pca2 <- dudi.pca(deug$tab, center = TRUE, scale = TRUE, scan = FALSE)
```




The script can be directly usd in Rstudio and R via the R package reticulate (https://rstudio.github.io/reticulate/).

```{r}
write.csv(deug$tab,file="data/deugtab.csv")
```



## Data importation

see for example: https://www.kaggle.com/code/arnopub/pandas-pr-sentation-des-dataframe

```{python}
import numpy as np 
import pandas as pd
deugtab = pd.read_csv('data/deugtab.csv')
deugtab
del deugtab['Unnamed: 0']
deugtab
```


# PCA from scratch in python
<a href="#top">Back to top</a>

***


PCA from scratch (https://towardsdatascience.com/principal-component-analysis-from-scratch-in-numpy-61843da1f967)

```{python}
# centering = TRUE
X= deugtab - deugtab.mean()
# Normalize
Z = X / X.std(ddof=0)
print('MEAN:')
print(Z.mean())
print('---'*15)
print('STD:')
print(Z.std(ddof=0))
```

diagonalisation and eigenvectors
```{python}
import numpy as np
len(Z)
ZZ = np.dot(Z.T, Z)/len(Z)
eigenvalues, eigenvectors = np.linalg.eig(ZZ)
D = np.diag(eigenvalues)
P = eigenvectors
Z_new = np.dot(Z, P)
```
valeur propres non ordonnées !!!!

Calculate the proportion of variance explained by each feature
```{python}
sum_eigenvalues = np.sum(eigenvalues)
sum_eigenvalues
prop_var = [i/sum_eigenvalues for i in eigenvalues]
```

Calculate the cumulative variance

```{python}
cum_var = [np.sum(prop_var[:i+1]) for i in range(len(prop_var))]
```

Plot scree plot from PCA
```{python}
import matplotlib.pyplot as plt
x_labels = ['PC{}'.format(i+1) for i in range(len(prop_var))]
plt.plot(x_labels, prop_var, marker='o', markersize=6, color='skyblue', linewidth=2, label='Proportion of variance')
plt.plot(x_labels, cum_var, marker='o', color='orange', linewidth=2, label="Cumulative variance")
plt.legend()
plt.title('Scree plot')
plt.xlabel('Principal components')
plt.ylabel('Proportion of variance')
plt.show()
```


# Collaboration R and python
<a href="#top">Back to top</a>

***

https://rstudio.github.io/reticulate/


```{r fig.width=8,fig.height=8}
library(reticulate)
library(ade4)
P <- py$P
colnames(P) <- paste("Axis",1:ncol(P),sep="")
rownames(P) <- colnames(py$deugtab)
par(mfrow=c(2,2))
s.corcircle(P,sub="Python version")
s.corcircle(pca2$c1,sub="R version")
plot(P[,1],pca2$c1[,1],panel.first=c(grid()),xlab="Python (axis 1)",ylab="R (axis 1)",pch=19);abline(0,-1,col="red")
plot(P[,2],pca2$c1[,2],panel.first=c(grid()),xlab="Python (axis 2)",ylab="R (axis 2)",pch=19);abline(0,-1,col="red")
```



```{r fig.width=8,fig.height=8}
coordli <- py$Z_new
colnames(coordli) <- paste("CS",1:ncol(coordli),sep="")
rownames(coordli) <- rownames(py$deugtab)
par(mfrow=c(2,2))
s.label(coordli,sub="Python version")
s.label(pca2$li,sub="R version")
plot(coordli[,1],pca2$li[,1],panel.first=c(grid()),xlab="Python (CS 1)",ylab="R (CS 1)",pch=19);abline(0,-1,col="red")
plot(coordli[,2],pca2$li[,2],panel.first=c(grid()),xlab="Python (CS 2)",ylab="R (CS 2)",pch=19);abline(0,-1,col="red")
```

```{r fig.width=10,fig.height=5}
par(mfrow=c(1,2))
plot(py$eigenvalues,pca2$eig,type="b",panel.first=c(grid()),pch=19); abline(0,1,col="red")
plot(sort(py$eigenvalues,decreasing = TRUE),pca2$eig,type="b",panel.first=c(grid()),pch=19); abline(0,1,col="red")
```

problem in the order of the eigenvalues !!!


```{r}
pca2$eig
py$D
py$eigenvalues
```

problem !!!

test of the pca from scikit-learn
```{python}
import sklearn.decomposition as sd
from sklearn.decomposition import PCA
pca = PCA(n_components=9)
Z2 = Z/np.sqrt(104)
pca.fit(Z2)
print(pca.explained_variance_ratio_)
print(pca.singular_values_)
print(pca.singular_values_*pca.singular_values_)
```


# Construction of the function "pydudi" (first prototype)
<a href="#top">Back to top</a>

***

based on the duality diagram (see more details below)

 * https://pbil.univ-lyon1.fr/R/pdf/tdr61.pdf
 * https://pbil.univ-lyon1.fr/R/pdf/stage3.pdf
 * https://pbil.univ-lyon1.fr/R/pdf/bs8.pdf


first test => need to adjust the weighting (test with COA)

```{python}
import os
import string
import re
import pandas as pd
import numpy as np

def pydudi(X,cw,lw,nf):
  dim = X.shape
  n = dim[0]
  p = dim[1]
  nf0 = nf-1
  # n=len(X)
  # p=len(X.columns)
  D = np.diag(np.sqrt(lw))
  Q = np.diag(np.sqrt(cw))
  
  # XtDXQ => problem  with Q !!!
  
  XD = np.dot(X.T,D).T
  XD = np.dot(XD,Q)
  XtX = np.dot(XD.T,XD)
  
  
  # decomposition
  eigenvalues, eigenvectors = np.linalg.eig(XtX)
  index = np.argsort(eigenvalues)[::-1]
  
  # np.nonzero(eigenvalues)[0]
  eigenvalues = eigenvalues.real
  eigenvectors = eigenvectors.real
  eigenvalues=eigenvalues[index]
  eigenvectors=eigenvectors[:,index]
  # results
  rank = len(np.nonzero(eigenvalues)[0])
  C1 = np.dot(np.diag(1/np.sqrt(cw)),eigenvectors[:,0:nf])
  #C1 = eigenvectors[:,0:nf]
  XQ = np.dot(X,np.diag(cw))
  Li = np.dot(XQ, C1)
  
  # need to adjust the weighting (problem with sqrt)
  L1 = np.dot(Li,np.diag(1/np.sqrt(eigenvalues[0:nf])))
  Co = np.dot(C1,np.diag(np.sqrt(eigenvalues[0:nf])))
  return eigenvalues,rank,Li,L1,Co,C1,nf;

###
```

## Test pydudi with PCA

```{python}
import numpy as np 
import pandas as pd
deugtab = pd.read_csv('data/deugtab.csv')
deugtab
del deugtab['Unnamed: 0']
deugtab
X= deugtab - deugtab.mean()
# Normalize
X = X / X.std(ddof=0)
dim = X.shape
n = dim[0]
p = dim[1]
# lw = pd.DataFrame(np.repeat(1/len(X),len(X)))[0]
# cw = pd.DataFrame(np.repeat(1,len(X.columns)))[0]
lw = pd.DataFrame(np.repeat(1/n,n))[0]
cw = pd.DataFrame(np.repeat(1,p))[0]
ted = pydudi(X,cw,lw,2)
```



```{r}
library(reticulate)
names(py$ted) <- c("eig","rank","li","l1","co","c1","nf")
coordli <- py$ted$li
par(mfrow=c(2,2))
s.label(coordli,sub="Python version")
s.label(pca2$li,sub="R version")
plot(coordli[,1],pca2$li[,1],panel.first=c(grid()),xlab="Python (CS 1)",ylab="R (CS 1)",pch=19);abline(0,-1,col="red")
plot(coordli[,2],pca2$li[,2],panel.first=c(grid()),xlab="Python (CS 2)",ylab="R (CS 2)",pch=19);abline(0,-1,col="red")
```

```{r}
coordl1 <- py$ted$l1
par(mfrow=c(2,2))
s.label(coordl1,sub="Python version")
s.label(pca2$l1,sub="R version")
plot(coordl1[,1],pca2$l1[,1],panel.first=c(grid()),xlab="Python (CS 1)",ylab="R (CS 1)",pch=19);abline(0,-1,col="red")
plot(coordl1[,2],pca2$l1[,2],panel.first=c(grid()),xlab="Python (CS 2)",ylab="R (CS 2)",pch=19);abline(0,-1,col="red")
```

```{r fig.height=10,fig.width=10}
par(mfrow=c(2,2))
s.corcircle(py$ted$co,sub="Python version")
s.corcircle(pca2$co,sub="R version")
plot(py$ted$c1[,1],pca2$c1[,1],panel.first=c(grid()),xlab="Python (axis 1)",ylab="R (axis 1)",pch=19);abline(0,-1,col="red")
plot(py$ted$c1[,2],pca2$c1[,2],panel.first=c(grid()),xlab="Python (axis 2)",ylab="R (axis 2)",pch=19);abline(0,-1,col="red")
```

## Test pydudi with COA

COA: correspondence analysis.

     Benzécri, J.P. and Coll. (1973) _L'analyse des données. II
     L'analyse des correspondances_, Bordas, Paris. 1-620.
     
     Greenacre, M. J. (1984) _Theory and applications of correspondence
     analysis_, Academic Press, London.


Si R help from ade4 (dudi.coa) and https://pbil.univ-lyon1.fr/R/pdf/stage4.pdf


```{r}
data(rpjdl)
chisq.test(rpjdl$fau)$statistic
rpjdl.coa <- coa1 <- dudi.coa(rpjdl$fau, scannf = FALSE, nf = 4)
sum(rpjdl.coa$eig)*rpjdl.coa$N # the same
```

```{r}
write.csv(rpjdl$fau,file="data/fau.csv")
```



```{python}
# import numpy as np 
import pandas as pd
fau = pd.read_csv('data/fau.csv')
fau
del fau['Unnamed: 0']
fau
```



```{python}
import numpy as np

X=fau
sumX = fau.sum().sum()
sumCol =  fau.sum(axis=0)
sumRow = fau.sum(axis=1)

pij = X/sumX
pi = sumRow/sumX
pj = sumCol/sumX

Dj = np.diag(1/pj)
Di = np.diag(1/pi)

Z = np.dot(Di,pij)
Z= np.dot(Z,Dj)
Z.shape
Z =Z - 1
Z = np.nan_to_num(Z)

# Normalize
lw = pi
cw = pj
D= np.diag(np.sqrt(pi))
Q= np.diag(np.sqrt(pj))
X =Z
ted = pydudi(Z,cw,lw,2)
```


```{r}
require(reticulate)
names(py$ted) <- c("eig","rank","li","l1","co","c1","nf")
```



```{r}
head(t(t(py$X)%*%py$D))[,1]
head(py$X*diag(py$D))[,1]
```


```{r}
XD <- t(t(py$X)%*%py$D)
head(XD%*%py$Q)[,1]
head(sweep(XD,2,diag(py$Q),"*"))[,1]
```

```{r}
coa1$eig
py$ted$eig
```


ok for the eigenvalues, probleme dans le calcul des coordonnées ???

```{r}
coordli <- py$ted$li
par(mfrow=c(2,2))
s.label(coordli,sub="Python version")
s.label(coa1$li,sub="R version")
plot(coordli[,1],coa1$li[,1],panel.first=c(grid()),xlab="Python (CS 1)",ylab="R (CS 1)",pch=19);abline(0,-1,col="red")
plot(coordli[,2],coa1$li[,2],panel.first=c(grid()),xlab="Python (CS 2)",ylab="R (CS 2)",pch=19);abline(0,-1,col="red")
```

```{r}
coordl1 <- py$ted$l1
par(mfrow=c(2,2))
s.label(coordl1,sub="Python version")
s.label(coa1$l1,sub="R version")
plot(coordl1[,1],coa1$l1[,1],panel.first=c(grid()),xlab="Python (CS 1)",ylab="R (CS 1)",pch=19);abline(0,-1,col="red")
plot(coordl1[,2],coa1$l1[,2],panel.first=c(grid()),xlab="Python (CS 2)",ylab="R (CS 2)",pch=19);abline(0,-1,col="red")
```

```{r fig.height=10,fig.width=10}
par(mfrow=c(2,2))
plot(py$Z[,1],coa1$tab[,1],panel.first=c(grid()),xlab="Python (col 1 of Z)",ylab="R (col 1 of tab)",pch=19);abline(0,1,col="red")
plot(py$ted$c1[,1],coa1$c1[,1],panel.first=c(grid()),xlab="Python (normed Axis 1)",ylab="R (normed Axis 1)",pch=19);abline(0,-1,col="red")
plot(py$ted$co[,1],coa1$co[,1],panel.first=c(grid()),xlab="Python (Axis 1)",ylab="R (Axis 1)",pch=19);abline(0,-1,col="red")
```


=> need to check the weight !!! => Ok !!

## Test pydudi with MCA

MCA: Multiple Correspondence Analysis

    Tenenhaus, M. & Young, F.W. (1985) An analysis and synthesis of multiple correspondence analysis, optimal scaling, dual scaling, homogeneity analysis ans other     methods for quantifying categorical multivariate data. Psychometrika, 50, 1, 91-119.
    Lebart, L., A. Morineau, and M. Piron. 1995. Statistique exploratoire multidimensionnelle. Dunod, Paris.


Si R help from ade4 (dudi.acm) and https://pbil.univ-lyon1.fr/R/pdf/stage4.pdf. the description of the methods is given below:

 - https://pbil.univ-lyon1.fr/R/pdf/tdr521.pdf
 - https://pbil.univ-lyon1.fr/R/pdf/tdr52.pdf

duality diagram



```{r}
data(ours)
summary(ours)
acm1 <- dudi.acm(ours, scan = FALSE)
```


```{r}
write.csv(ours,file="data/ours.csv")
```


importation with pandas
```{python}
# import numpy as np 
import pandas as pd
ours = pd.read_csv('data/ours.csv')
ours
del ours['Unnamed: 0']
ours
```

```{python}
def disjonctif(X):
  X_cat = X.astype("category")
  X_dis = pd.get_dummies(X_cat)
  X_dis = X_dis*1
  return X_dis ;
```


preparation of the triplet
```{python}
Xdis = disjonctif(ours)
m = Xdis.shape[1]
n = Xdis.shape[0]
v = ours.shape[1]
lw = pd.DataFrame(np.repeat(1/n,n))[0]
D = np.diag(lw)
cw = np.dot(np.dot(Xdis.T,D), np.ones(n)) 
Dm = np.diag(cw)
X = np.dot(Xdis,np.diag(1/cw))-1
cw = cw/v
ted = pydudi(X,cw,lw,2)
```



```{r}
require(reticulate)
names(py$ted) <- c("eig","rank","li","l1","co","c1","nf")
```


```{r}
coordli <- py$ted$li
par(mfrow=c(2,2))
s.label(coordli,sub="Python version")
s.label(acm1$li,sub="R version")
plot(coordli[,1],acm1$li[,1],panel.first=c(grid()),xlab="Python (CS 1)",ylab="R (CS 1)",pch=19);abline(0,1,col="red")
plot(coordli[,2],acm1$li[,2],panel.first=c(grid()),xlab="Python (CS 2)",ylab="R (CS 2)",pch=19);abline(0,-1,col="red")
```

```{r}
coordl1 <- py$ted$l1
par(mfrow=c(2,2))
s.label(coordl1,sub="Python version")
s.label(acm1$l1,sub="R version")
plot(coordl1[,1],acm1$l1[,1],panel.first=c(grid()),xlab="Python (CS 1)",ylab="R (CS 1)",pch=19);abline(0,1,col="red")
plot(coordl1[,2],acm1$l1[,2],panel.first=c(grid()),xlab="Python (CS 2)",ylab="R (CS 2)",pch=19);abline(0,-1,col="red")
```

```{r fig.height=10,fig.width=10}
par(mfrow=c(2,2))
plot(py$X[,1],acm1$tab[,1],panel.first=c(grid()),xlab="Python (col 1 of Z)",ylab="R (col 1 of tab)",pch=19);abline(0,1,col="red")
plot(py$ted$c1[,1],acm1$c1[,1],panel.first=c(grid()),xlab="Python (normed Axis 1)",ylab="R (normed Axis 1)",pch=19);abline(0,1,col="red")
plot(py$ted$co[,1],acm1$co[,1],panel.first=c(grid()),xlab="Python (Axis 1)",ylab="R (Axis 1)",pch=19);abline(0,1,col="red")
```


# Class and object dudi
<a href="#top">Back to top</a>

***


define the structure of the object and description of the elements based on object 'dudi' from R package ade4 (and ADE-4):

 - tab
 - eig
 - rank
 - c1
 - co
 - l1
 - li

heritage for the PCA and COA (and other methods):

 - tab
 - lw
 - cw
 - dudi
 - and specific elements
 
## Object **Dudi**

```{python}
class Dudi:
  def __init__(self,tab,cw,lw,eig,rank,nf,c1,co,l1,li):
    self.tab = tab
    self.cw = cw
    self.lw = lw
    self.eig = eig
    self.rank = rank
    self.nf = nf
    self.c1 = c1
    self.co = co
    self.l1 = l1
    self.li = li

def pyDudi(X,cw,lw,nf):
  dim = X.shape
  n = dim[0]
  p = dim[1]
  nf0 = nf-1
  # n=len(X)
  # p=len(X.columns)
  D = np.diag(np.sqrt(lw))
  Q = np.diag(np.sqrt(cw))
  
  # XtDXQ => problem  with Q !!!
  XD = np.dot(X.T,D).T
  XD = np.dot(XD,Q)
  XtX = np.dot(XD.T,XD)
  
  # decomposition
  eigenvalues, eigenvectors = np.linalg.eig(XtX)
  index = np.argsort(eigenvalues)[::-1]
  
  # np.nonzero(eigenvalues)[0]
  eigenvalues = eigenvalues.real
  eigenvectors = eigenvectors.real
  eigenvalues=eigenvalues[index]
  eigenvectors=eigenvectors[:,index]
  # results
  rank = len(np.nonzero(eigenvalues)[0])
  C1 = np.dot(np.diag(1/np.sqrt(cw)),eigenvectors[:,0:nf])
  #C1 = eigenvectors[:,0:nf]
  XQ = np.dot(X,np.diag(cw))
  Li = np.dot(XQ, C1)
  
  # need to adjust the weighting (problem with sqrt)
  L1 = np.dot(Li,np.diag(1/np.sqrt(eigenvalues[0:nf])))
  Co = np.dot(C1,np.diag(np.sqrt(eigenvalues[0:nf])))
  return Dudi(X,cw,lw,eigenvalues,rank,nf,C1,Co,L1,Li);
```

## test pour PCA
 
 

 
 
 
```{python}
def pyPCA(X,cw=None,lw=None,nf=2,center=True,scale=True):
  dim = X.shape
  n = dim[0]
  p = dim[1]  
  if center:
    X = X-X.mean()
  if scale:
    X = X/X.std(ddof=0)
  if lw==None :
    lw = pd.DataFrame(np.repeat(1/n,n))[0]
  if cw==None :
    cw = pd.DataFrame(np.repeat(1,p))[0]
  dudi = pyDudi(X,cw,lw,nf)  
  return dudi;
```


```{python}
pca1 = pyPCA(deugtab)
```

## test pour COA
 
```{python}
def pyCOA(X,nf=2):
  sumX = X.sum().sum()
  sumCol =  X.sum(axis=0)
  sumRow = X.sum(axis=1)
  pij = X/sumX
  pi = sumRow/sumX
  pj = sumCol/sumX
  Dj = np.diag(1/pj)
  Di = np.diag(1/pi)
  Z = np.dot(Di,pij)
  Z = np.dot(Z,Dj)
  Z = Z - 1
  Z = np.nan_to_num(Z)
  dudi = pyDudi(Z,pj,pi,nf)  
  return dudi;
```


```{python}
coa1 = pyCOA(fau)
```

## test pour ACM
 
```{python}
def disjonctif(X):
  X_cat = X.astype("category")
  X_dis = pd.get_dummies(X_cat)
  X_dis = X_dis*1
  return X_dis;

def pyACM(X,lw=None,nf=2):
  Xdis = disjonctif(ours)
  m = Xdis.shape[1]
  n = Xdis.shape[0]
  v = ours.shape[1]
  if lw==None :
    lw = pd.DataFrame(np.repeat(1/n,n))[0]
  D = np.diag(lw)
  cw = np.dot(np.dot(Xdis.T,D), np.ones(n)) 
  Dm = np.diag(cw)
  X = np.dot(Xdis,np.diag(1/cw))-1
  cw = cw/v
  dudi = pyDudi(X,cw,lw,nf)    
  return dudi;
```


```{python}
mca1 = pyACM(ours)
```




# Construction of the function dudi in Rcpp


## Test Rcpp: correlation between two variables

```{r}
require(parallel)
require(Rcpp)
require(RcppArmadillo)
#Sys.setenv("PKG_CXXFLAGS"="-fopenmp")
#Sys.setenv("PKG_LIBS"="-fopenmp")
sourceCpp(file.path(getwd(),"src","utility.cpp"))
```

correlation coefficient

```{r}
cor(1:10, 2:11) 
CORR(1:10, 2:11) 
```

computation of the area under curve
```{r}
```




## Object dudi

dudi obect with Rcpparmadillo

```{r}
sourceCpp(file.path(getwd(),"src","utility.cpp"))
pca2 <- dudi.pca(deug$tab, center = TRUE, scale = TRUE, scan = FALSE)
test <- arc_dudi(as.matrix(pca2$tab),pca2$cw,pca2$lw,2)
test$eig
pca2$eig
```



```{r}
coordli <- test$li
par(mfrow=c(2,2))
s.label(coordli,sub="RcppArmadillo version")
s.label(pca2$li,sub="R version")
plot(coordli[,1],pca2$li[,1],panel.first=c(grid()),xlab="RcppArmadillo (CS 1)",ylab="R (CS 1)",pch=19);abline(0,-1,col="red")
plot(coordli[,2],pca2$li[,2],panel.first=c(grid()),xlab="RcppArmadillo (CS 2)",ylab="R (CS 2)",pch=19);abline(0,-1,col="red")
```

```{r}
coordl1 <- test$l1
par(mfrow=c(2,2))
s.label(coordl1,sub="RcppArmadillo version")
s.label(pca2$l1,sub="R version")
plot(coordl1[,1],pca2$l1[,1],panel.first=c(grid()),xlab="RcppArmadillo (CS 1)",ylab="R (CS 1)",pch=19);abline(0,-1,col="red")
plot(coordl1[,2],pca2$l1[,2],panel.first=c(grid()),xlab="RcppArmadillo (CS 2)",ylab="R (CS 2)",pch=19);abline(0,-1,col="red")
```

```{r fig.height=10,fig.width=10}
par(mfrow=c(2,2))
s.corcircle(test$co,sub="RcppArmadillo version")
s.corcircle(pca2$co,sub="R version")
plot(test$c1[,1],pca2$c1[,1],panel.first=c(grid()),xlab="RcppArmadillo (axis 1)",ylab="R (axis 1)",pch=19);abline(0,-1,col="red")
plot(test$c1[,2],pca2$c1[,2],panel.first=c(grid()),xlab="RcppArmadillo (axis 2)",ylab="R (axis 2)",pch=19);abline(0,-1,col="red")
```


## Test PCA

argument for the function cpca:
 
 * X: matrix
 * cw: numeric vector corresponding to the column weighting
 * lw: numeric vector corresponding to the row weighting
 * nf: number of selected axes
 * center: 1 for TRUE, 0 for FALSE
 * scale: 1 for TRUE, 0 for FALSE 


```{r}
sourceCpp(file.path(getwd(),"src","utility.cpp"))
test2 <- arc_pca(as.matrix(deug$tab),pca2$cw,pca2$lw,2,1,1)
pca2$eig
test2$eig
```


```{r}
coordli <- test2$li
par(mfrow=c(2,2))
s.label(coordli,sub="RcppArmadillo version")
s.label(pca2$li,sub="R version")
plot(coordli[,1],pca2$li[,1],panel.first=c(grid()),xlab="RcppArmadillo (CS 1)",ylab="R (CS 1)",pch=19);abline(0,-1,col="red")
plot(coordli[,2],pca2$li[,2],panel.first=c(grid()),xlab="RcppArmadillo (CS 2)",ylab="R (CS 2)",pch=19);abline(0,-1,col="red")
```

```{r}
coordl1 <- test2$l1
par(mfrow=c(2,2))
s.label(coordl1,sub="RcppArmadillo version")
s.label(pca2$l1,sub="R version")
plot(coordl1[,1],pca2$l1[,1],panel.first=c(grid()),xlab="RcppArmadillo (CS 1)",ylab="R (CS 1)",pch=19);abline(0,-1,col="red")
plot(coordl1[,2],pca2$l1[,2],panel.first=c(grid()),xlab="RcppArmadillo (CS 2)",ylab="R (CS 2)",pch=19);abline(0,-1,col="red")
```

```{r fig.height=10,fig.width=10}
par(mfrow=c(2,2))
s.corcircle(test2$co,sub="RcppArmadillo version")
s.corcircle(pca2$co,sub="R version")
plot(test2$c1[,1],pca2$c1[,1],panel.first=c(grid()),xlab="RcppArmadillo (axis 1)",ylab="R (axis 1)",pch=19);abline(0,-1,col="red")
plot(test2$c1[,2],pca2$c1[,2],panel.first=c(grid()),xlab="RcppArmadillo (axis 2)",ylab="R (axis 2)",pch=19);abline(0,-1,col="red")
```




## Test COA


```{r}
sourceCpp(file.path(getwd(),"src","utility.cpp"))
data(rpjdl)
chisq.test(rpjdl$fau)$statistic
rpjdl.coa <- coa1 <- dudi.coa(rpjdl$fau, scannf = FALSE, nf = 4)
sum(rpjdl.coa$eig)*rpjdl.coa$N # the same
test <- arc_coa(as.matrix(rpjdl$fau),nf=2)
```




```{r}
coordli <- test$li
par(mfrow=c(2,2))
s.label(coordli,sub="Rcpp-Armadillo version")
s.label(coa1$li,sub="R version")
plot(coordli[,1],coa1$li[,1],panel.first=c(grid()),xlab="Rcpp-Armadillo (CS 1)",ylab="R (CS 1)",pch=19);abline(0,1,col="red")
plot(coordli[,2],coa1$li[,2],panel.first=c(grid()),xlab="Rcpp-Armadillo (CS 2)",ylab="R (CS 2)",pch=19);abline(0,-1,col="red")
```

```{r}
coordl1 <- test$l1
par(mfrow=c(2,2))
s.label(coordl1,sub="Rcpp-Armadillo version")
s.label(coa1$l1,sub="R version")
plot(coordl1[,1],coa1$l1[,1],panel.first=c(grid()),xlab="Rcpp-Armadillo (CS 1)",ylab="R (CS 1)",pch=19);abline(0,1,col="red")
plot(coordl1[,2],coa1$l1[,2],panel.first=c(grid()),xlab="Rcpp-Armadillo (CS 2)",ylab="R (CS 2)",pch=19);abline(0,-1,col="red")
```

```{r fig.height=10,fig.width=10}
par(mfrow=c(2,2))
plot(test$X[,1],coa1$tab[,1],panel.first=c(grid()),xlab="Rcpp-Armadillo (col 1 of Z)",ylab="R (col 1 of tab)",pch=19);abline(0,1,col="red")
plot(test$c1[,1],coa1$c1[,1],panel.first=c(grid()),xlab="Rcpp-Armadillo (normed Axis 1)",ylab="R (normed Axis 1)",pch=19);abline(0,1,col="red")
plot(test$co[,1],coa1$co[,1],panel.first=c(grid()),xlab="Rcpp-Armadillo (Axis 1)",ylab="R (Axis 1)",pch=19);abline(0,1,col="red")
```





## Test MCA

```{r}
data(ours)
summary(ours)
acm1 <- dudi.acm(ours, scan = FALSE)
```


for the first version of mca, we give the **disjonctif** table and the variable number.
```{r}
sourceCpp(file.path(getwd(),"src","utility.cpp"))
test <- arc_mca_proto(as.matrix(acm.disjonctif(ours)),lw=acm1$lw,v=ncol(ours),nf=2)
```


```{r}
coordli <- test$li
par(mfrow=c(2,2))
s.label(coordli,sub="Rcpp-Armadillo version")
s.label(acm1$li,sub="R version")
plot(coordli[,1],acm1$li[,1],panel.first=c(grid()),xlab="Rcpp-Armadillo (CS 1)",ylab="R (CS 1)",pch=19);abline(0,-1,col="red")
plot(coordli[,2],acm1$li[,2],panel.first=c(grid()),xlab="Rcpp-Armadillo (CS 2)",ylab="R (CS 2)",pch=19);abline(0,1,col="red")
```

```{r}
coordl1 <- test$l1
par(mfrow=c(2,2))
s.label(coordl1,sub="Python version")
s.label(acm1$l1,sub="R version")
plot(coordl1[,1],acm1$l1[,1],panel.first=c(grid()),xlab="Rcpp-Armadillo (CS 1)",ylab="R (CS 1)",pch=19);abline(0,-1,col="red")
plot(coordl1[,2],acm1$l1[,2],panel.first=c(grid()),xlab="Rcpp-Armadillo (CS 2)",ylab="R (CS 2)",pch=19);abline(0,1,col="red")
```

```{r fig.height=10,fig.width=10}
par(mfrow=c(2,2))
plot(test$X[,1],acm1$tab[,1],panel.first=c(grid()),xlab="Rcpp-Armadillo (col 1 of Z)",ylab="R (col 1 of tab)",pch=19);abline(0,1,col="red")
plot(test$c1[,1],acm1$c1[,1],panel.first=c(grid()),xlab="Rcpp-Armadillo (normed Axis 1)",ylab="R (normed Axis 1)",pch=19);abline(0,-1,col="red")
plot(test$co[,1],acm1$co[,1],panel.first=c(grid()),xlab="Rcpp-Armadillo (Axis 1)",ylab="R (Axis 1)",pch=19);abline(0,-1,col="red")
```

encoding **disjonctif** table
```{r}
sourceCpp(file.path(getwd(),"src","utility.cpp"))
w <- factor(c("a","a","b","c"))
acm.util(w)
arc_acmutil(w)
w1 <- factor(c("e","d","e","f"))
arc_disjonctif(cbind(w,w1))
```

```{r}
test <- arc_mca( as.matrix(apply(ours,2,function(x) as.numeric(factor(x)))),lw=acm1$lw,nf=2)
```


```{r}
coordli <- test$li
par(mfrow=c(2,2))
s.label(coordli,sub="Rcpp-Armadillo version")
s.label(acm1$li,sub="R version")
plot(coordli[,1],acm1$li[,1],panel.first=c(grid()),xlab="Rcpp-Armadillo (CS 1)",ylab="R (CS 1)",pch=19);abline(0,-1,col="red")
plot(coordli[,2],acm1$li[,2],panel.first=c(grid()),xlab="Rcpp-Armadillo (CS 2)",ylab="R (CS 2)",pch=19);abline(0,1,col="red")
```

```{r}
coordl1 <- test$l1
par(mfrow=c(2,2))
s.label(coordl1,sub="Python version")
s.label(acm1$l1,sub="R version")
plot(coordl1[,1],acm1$l1[,1],panel.first=c(grid()),xlab="Rcpp-Armadillo (CS 1)",ylab="R (CS 1)",pch=19);abline(0,-1,col="red")
plot(coordl1[,2],acm1$l1[,2],panel.first=c(grid()),xlab="Rcpp-Armadillo (CS 2)",ylab="R (CS 2)",pch=19);abline(0,1,col="red")
```

```{r fig.height=10,fig.width=10}
par(mfrow=c(2,2))
plot(test$X[,1],acm1$tab[,1],panel.first=c(grid()),xlab="Rcpp-Armadillo (col 1 of Z)",ylab="R (col 1 of tab)",pch=19);abline(0,1,col="red")
plot(test$c1[,1],acm1$c1[,1],panel.first=c(grid()),xlab="Rcpp-Armadillo (normed Axis 1)",ylab="R (normed Axis 1)",pch=19);abline(0,-1,col="red")
plot(test$co[,1],acm1$co[,1],panel.first=c(grid()),xlab="Rcpp-Armadillo (Axis 1)",ylab="R (Axis 1)",pch=19);abline(0,-1,col="red")
```



# References
<a href="#top">Back to top</a>

***

-   R <http://www.R-project.org/>.
-   R Core Team (2013). R: A language and environment for statistical
    computing. R Foundation for Statistical Computing, Vienna, Austria.
    URL <http://www.R-project.org/>.
-   Dray S, Dufour A (2007). “The ade4 Package: Implementing the Duality Diagram for Ecologists.” _Journal of Statistical Software_, *22*(4), 1-20. doi:10.18637/jss.v022.i04
  <https://doi.org/10.18637/jss.v022.i04>.
-   Chessel D, Dufour A, Thioulouse J (2004). “The ade4 Package - I: One-Table Methods.” _R News_, *4*(1), 5-10. <https://cran.r-project.org/doc/Rnews/>.
-   Dray S, Dufour A, Chessel D (2007). “The ade4 Package - II: Two-Table and K-Table Methods.” _R News_, *7*(2), 47-52. <https://cran.r-project.org/doc/Rnews/>.
-   Thioulouse J, Dray S, Dufour A, Siberchicot A, Jombart T, Pavoine S (2018). _Multivariate Analysis of Ecological Data with ade4_. Springer. doi:10.1007/978-1-4939-8850-1 <https://doi.org/10.1007/978-1-4939-8850-1>.
-   Escoufier, Y. (1987) The duality diagram : a means of better practical applications In Development in numerical ecology, Legendre, P. & Legendre, L. (Eds.) NATO advanced Institute, Serie G. Springer Verlag, Berlin, 139–156.
-   Benzécri, J.P. and Coll. (1973) _L'analyse des données. II L'analyse des correspondances_, Bordas, Paris. 1-620.
-   Greenacre, M. J. (1984) _Theory and applications of correspondence analysis_, Academic Press, London.
-   Tenenhaus, M. & Young, F.W. (1985) An analysis and synthesis of multiple correspondence analysis, optimal scaling, dual scaling, homogeneity analysis ans other methods for quantifying categorical multivariate data. Psychometrika, 50, 1, 91-119.
-   Lebart, L., A. Morineau, and M. Piron. 1995. Statistique exploratoire multidimensionnelle. Dunod, Paris.


# Appendix
<a href="#top">Back to top</a>

***


## Additional R functions

```{r}
source("/export/scratch/GITprojects/pbtools/trunk/Rcode/Rgraphics-0.1.R")
```

## Session information

```{r}
print(sessionInfo(),locale=FALSE)
```

